---
title: "Map Algebra for Site Selection in R "
author: "Kwesie Benjamin"
date: "5/12/2021"
output:
  rmdformats::readthedown:
    gallery: yes
    highlight: tango
    lightbox: yes
    self_contained: yes
    thumbnails: yes
  html_document:
    df_print: paged
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment="",collapse = TRUE, message = FALSE,warning = FALSE)
```


## **Introduction**

The Department of Planning of the (imaginary) oasis Aïn Kju Dzjis is planning to build the National
Institute for Oasis Management (NIOM) in their village. They came up with the following conditions 
for building the complex:

1. No industry, mine or landfill within 300 meters of the new complex;
2. Not on locations presently in use for buildings, water or roads;
3. The slope should be less than or equal to 3%;
4. The distance from an existing road should be less than 500 meters;
5. The area should be contiguous;
6. The area should be greater than, or equal to 2 hectares.

They hired you as a GIS consultant to find the most appropriate locations. You will use map algebra 
to perform the required multicriteria analysis.`(Van der Kwast, 2018)`.


Install and load packages as listed below.
```{r message=FALSE,warning=FALSE,eval=FALSE}
install.packages("raster",dependencies = T)
install.packages("tmap")
install.packages("tidyverse")
install.packages("sf")
install.packages("units")
```




Now we load the libraries. For more information on packages use `help()`  (eg., `help("raster")`).
```{r message=FALSE, warning=FALSE}
#Do well to set the working directory
setwd("C:/MapAlgebra")

library(raster)
library(tmap)
library(tidyverse)
library(sf)
library(units)
library(mapview)
```




Loading of Data  

* Functions to be used in this section:
    + **raster {raster}**
    + **tmap {tmap}** 

To begin, we laod in all tif files`(.tif)` using the function `raster()`.
```{r message=FALSE, warning=FALSE}
buildings <- raster("building.tif")
isroad <- raster("isroad.tif")
iswater <- raster("iswater.tif")
roads <- raster("roads.tif")
topo <- raster("topo.tif")
```




Lets check the properties of the RasterLayer Object using `print()`.
```{r message=FALSE, warning=FALSE}
cat("buildings\n"); print(buildings) ; cat("isroad\n"); print(isroad); cat("iswater\n"); print(iswater) ;
cat("roads\n"); print(roads) ; cat("topo\n"); print(topo)
```
It is very clear all raster layer have the same spatial resolution and same UTM projection.





However, to be sure we can check this in R using the function `CompareRaster()`.
```{r message=FALSE, warning=FALSE}
compareRaster(buildings,isroad,iswater,roads, topo,
              crs = TRUE,# CRS of raster layers
              res = TRUE #Resolution of raster layers
              )
```




## **Condition 1**

No industry, mine or Landfill within 300 metres of the complex.
The raster layer building, is discreete with values;

* 0 - None
* 1 - House
* 2 - Public building
* 3 - landfill
* 4 - Industry
* 5 - Mine

```{r message=FALSE, warning=FALSE}
unique(buildings) # It indicates the unique discrete values in the raster layer.
```



We need to re-classify the raster layer with;

* 0 - No industry
* 1 - Industry
In R we can do this using the functions `calc()` or `reclassify()`. However at this stage we would learn to use the `clac()` function.
```{r message=FALSE,warning=FALSE}
  IndustryClass<- calc(buildings,
                       fun = function(x){
                         ifelse(x>=3,1, 0) # We doing this in other to create a boolean(0,1).
                       })

summary(IndustryClass)
```



In R, the distance function `distance()` computes distance from NA cells to non-NA cells. Hence we need to assing NA to all  pixels with values of (0) in **industry_class**.
```{r message=FALSE, warning=FALSE}
IndustryClass[IndustryClass==0] = NA

unique(values(IndustryClass))
```



The IndustryDist layer is UTM projected with (+units = m), hence calculated distance would be in metres.
```{r message=FALSE, warning=FALSE}
IndustryDist <- distance(IndustryClass) 

summary(IndustryDist)
```





Now lets plot the **industry_dist** layer in `tmap()`.Areas close to the industry are shown in red.
```{r message=FALSE, warning=FALSE}
tm_shape(IndustryDist)+
  tm_raster(style='cont', palette = "RdYlGn", title = 'Distances to Industry') +
  tm_layout(legend.outside = TRUE)

```




Now we enforce the condition of **Industry,Mine or Landfill should be 300 metres** away from complex. 
```{r message=FALSE,warning=FALSE}
NoIndustry <- IndustryDist >= 300 # Output raster is now discrete and can be categorised.

tm_shape(NoIndustry)+
  tm_raster(style = 'cat', 
            title = " 300 metres away from industry",
            labels = c("No Industry", "Industry at 300m"),
            palette = c("red", "green"))+ 
  tm_layout(legend.outside = TRUE)
```




## **Condition 2** 

Not on locations presently in use for buildings, water or roads. Note that layers are mapped to values of 0(indicating the absence of buildings,Water-way,Roads).
```{r message=FALSE, warning=FALSE}
NoBuilding <- buildings == 0
NoWater <- iswater== 0
NoRoad <- isroad == 0
```




Lets style and Map the layers fufilling condition 2 above.
```{r message=FALSE, warning=FALSE}
NoBuilding_map <- tm_shape(NoBuilding)+
  tm_raster(style = 'cat',title= "", labels = c("Is Building", "No Building"),
            palette = c("red", "green"))+ tm_layout(main.title = "Building Map" )

NoWater_map <- tm_shape(NoWater)+
  tm_raster(style = 'cat',title="", labels = c("Is Water", "No Water"),
            palette = c("red", "green"))+ tm_layout(main.title = "Water Map")

NoRoad_map <- tm_shape(NoRoad)+
  tm_raster(style = 'cat',title="", labels = c("Is Road", "No Road"),
            palette = c("red", "green"))+ tm_layout(main.title = "Road Map")

tmap_arrange(NoBuilding_map,NoWater_map,NoRoad_map,ncol = 3, nrow = 1)
```




## **Condition 3** 

The slope should be less than or equal to 3 %.
In R we can compute the Slope using the `terrain()` function however this would be done in degrees.
```{r message=FALSE, warning=FALSE}
SlopeTopo <- terrain(topo, opt="slope",unit ="degrees")

#Now lets plot the slope of 3 degress.
tm_shape(SlopeTopo)+
  tm_raster(style = 'cont',
            title = "Slope in degrees",
            palette = "-RdYlGn")+
  tm_layout(legend.outside= TRUE ,legend.outside.position = "right")
```




Now lets enforce the criteria of 3(%) percent slope. Note, slope was calculated in degrees.

Percent of slope is determined by dividing the amount of elevation change by the amount of horizontal distance covered,
and then multiplied by **(100)**.
Knowing the percent slope we can find the slope in degrees using inverse tangent. 
$tan^{-1}$ of 3% slope would be approximately **1.72** degrees.
```{r message=FALSE,warning=FALSE}
SlopeNotSteep <- SlopeTopo<= 1.72

#Now lets plot discrete map of slope ( <= 1.72 dgreees, equivalent of 3% )

tm_shape(SlopeNotSteep)+
  tm_raster(style = 'cat', 
            labels = c("Steep", "Not Steep"),
                       title = "Slope less than or equals 300 m",
                       palette = c("red","green"))+
  tm_layout(legend.outside= TRUE ,legend.outside.position = "right")
```




## **Condition 4** 

The distance from an existing road should be less than 500 metres.
The road raster layer is discrete and has 3 classes with values;

* 0 - No road
* 1 - Dirt road
* 2 - Tarmac road

The `reclassify()` function would be use to reclassify values into **0(No roads)** 
and **1(Dirt and Tarmac roads)**.
```{r message=FALSE, warning=FALSE}
RclRoads <- reclassify(roads,
                       rcl = matrix(c(0,0,0, # from 0 to 0 ->0
                                      1,2,1),# from 1 to 2 ->1
                                    ncol = 3,byrow = TRUE))

# Lets now map the 0 values of rclRoads layer to NA to allow for distance computation.
RclRoads[RclRoads==0]<- NA #We could have equally set 0 values to NA within the reclassify matrix.

unique(values(RclRoads))
```



Lets now compute the distances for the reclassified road Map using the `distance()` function.
```{r message=FALSE, warning=FALSE}
RoadDist<- distance(RclRoads)  

tm_shape(RoadDist)+
  tm_raster(style = 'cont', 
            title ="Computed distances from road" ,
            palette = "-RdYlGn")+
  tm_layout(legend.outside= TRUE ,
            legend.outside.position = "right")
```





Lets enforce condition **4.The distance from an existing road should be less than 500 metres**
```{r message=FALSE, warning=FALSE}
LessThan500m <- RoadDist <= 500

#lets style the layer
tm_shape(LessThan500m)+
  tm_raster(style = "cat",
            title = "Distance less than or equals 500 m",
            labels = c("Greater than 500m","Less than 500m"),
            palette = c("red","green"))+
  tm_layout(legend.outside= TRUE ,
            legend.outside.position = "right")
```





Lets combine all the criteria. In R we would multiply all the boolean condition.
```{r message=FALSE, warning=FALSE}
Niomgo <- NoIndustry*NoBuilding*NoWater*NoRoad*SlopeNotSteep*LessThan500m

#Lets set all zero pxels to NA. Weare aonly interested in pixels with value =1(Boolen True).
Niomgo[Niomgo==0]<- NA

#Lets chechk to be sure all zero's are now NA leaving values of 1 only.
summary(Niomgo)


tm_shape(Niomgo)+
  tm_raster(style = "cat",
            title = "",
            labels = "True for all Conditions",
            palette = "green")+
  tm_layout(legend.outside= TRUE ,legend.outside.position = "right")
            
```

 


## **Condition 5**

The area should be contiguous(share common border).
Contiguity forms the concepts of neighbors and  how the pixels relate(interact) with other pixels.

At this point, we would recategorize the rasta values base on contiguity.We would only consider Vertical and Horizontal contiguity(Rook case).After clumping using `clump()` we then convert the raster cells to polygons using the `rasterToPolygons()` function.

At this stage we would be using the dplyr pipe ` %>% ` quite alot.
```{r message=FALSE, warning=FALSE}
NiomgoClump <- Niomgo %>%
  clump(directions= 4) %>% #dirction = 4(rook case)
  rasterToPolygons() %>% 
  st_as_sf() %>% 
  group_by(clumps) %>% 
  summarize() #This dissolves polygons with same clump value(clumps) into one region.
  
  

NiomgoArea <- NiomgoClump %>% 
  group_by(clumps) %>% 
  summarise(
    Area_poly = geometry %>% #We calculate Area from geometry  
      sf::st_zm() %>% #Drops  Z or M dimension from geometry
      sf::st_area() %>% round(2))


NiomgoSelected <- NiomgoArea %>% 
  filter(Area_poly >= units::set_units(20000, m^2) )# Since Area_poly has a unit of meter square(m^2)
                                             # we need to also set unit of 200000 to meters(m^2)

```




## **Selected Locations**


Lets now plot a map of final selected areas.
```{r message=FALSE, warning=FALSE}
#In order to plot NiomgoSelected as categorical in tmap, we need to convert the 'clumps(feature)'
#from double to character.

NiomgoSelected["clumps"]<- as.character(NiomgoSelected$clumps)

tm_shape(NiomgoSelected)+
  tm_polygons(style = "cat", col= "clumps", title= "Selected Sites")+
  tm_layout( legend.outside = TRUE)
```




Lets also plot the selected site in `mapView()` to have a visual statisfaction of real-time location.
```{r}
mapView(NiomgoSelected,
        map.types="Esri.WorldImagery")
```



<!--html_preserve-->
<br>
<body>
  <style type="text/css">
	      a:hover{
	          color: red;
	      }
	      .tab{
	        margin-left: 40px;
	        padding-top:0;
	      }
	        
	      p{
	        	color: #555555;
	        	font-family: Arial,sans-serif;
	        	font-size: 16px;
	        	font-weight: normal;
	      }
	 </style>
	<div>
	  <h2><center> Acknowledgement</center></h2>
		<p style="text-alignment: justify">
		    Sincere gratitude to Dr.Hans van der Kwast the author of the book Qgis and Open Data for Hydrological Applications 
		    and His in-depth tutorial on Qgis on his  
		    youtube page <a href="http://www.youtube.com/c/HansvanderKwast"  target="_blanck">HansvanderKwast</a>.
		    This resource seeks to simulate what was done in the eBook QGIS and Open Data for Hydrological 
		    Applications-Exercise Manual.(Exercise 4: Map Algebra), however using the R programming language. As a student 
		    myself, I believe this workflow would endow students with some requisite raster data analysis skills using the R
		    programing language.
		 </p>
	</div>

	  <br>
	  <br>
  
	<div>
	    
		 <h2>
		    <center><b>References</b></centre>
		 </h2>
		 <br>
		 <p>
		    Lovelace,R.,Nowosad,J.,and Muenchow,J.(2019) <em>GeoComputation</em> with R*.1st edn.
		 </P>
		 <p class="tab">
		    Bocca Raton:Chapman and Hall/CRC.
		 </p>
		 <br>
		 <p>
		    Van der Kwast, J. (2018) <em>Qgis and Open Data for Hydrlogical Applications-ExerciseManual</em>.
		 </P>
		 <p class="tab" >
		    version3.4.1b.ocw.un-ihe.org.
		 </p>
		 <br>
		 <p>
		    Van der Kwast,J.(2018) 'Data Exercise 4:Spatial planning Using map algebra'.
		 </p>
		 <p class="tab" >
		    Available at:<a href= "https://ocw.un-ihe.org/course/view.php?id=11&section=2" target="_blank" > 
		    https://ocw.un-ihe.org/course/view.php?id=11&section=2 </a> (Accessed April 2021). 
		 </p>
	  
	</div>
</body>

<!--/html_preserve-->

